---
title: "Module 5 Theory — TLF Tables (Summary Statistics)"
format: html
editor: visual
---

# 📊 Module 5 — TLF Tables: Summary Statistics

## 🎯 Learning Objectives

By the end of this module, you will:

-   Understand how to compute descriptive statistics by treatment arm
-   Learn to create summary tables with `tidyverse` and `gtsummary`
-   Create tables for continuous and categorical variables
-   Practice exporting tables to Word or HTML
-   Use Copilot to help build, format, and export summary tables

------------------------------------------------------------------------

## 📋 1. What Are Summary Tables?

Summary tables report aggregated statistics such as: - N, mean, standard deviation (SD), median, min, max - Frequency counts and percentages

They are commonly used to describe: - Baseline demographics (e.g., age, sex, race) - Safety/efficacy endpoints

------------------------------------------------------------------------

## 🧰 2. Key Packages for Summary Tables

| Package     | Purpose                            |
|-------------|------------------------------------|
| `dplyr`     | Data manipulation                  |
| `gtsummary` | Summary statistics tables          |
| `tidyverse` | Core tools for data wrangling      |
| `gt`        | Underlying engine for final output |

------------------------------------------------------------------------

## 📦 3. Example Data

```{r}
library(dplyr)
library(gtsummary)

demo <- tibble::tibble(
  USUBJID = paste0("01-", sprintf("%03d", 1:10)),
  AGE = c(34, 58, 45, 63, 39, 52, 60, 41, 50, 48),
  SEX = c("M", "F", "M", "F", "F", "M", "F", "M", "F", "F"),
  ARM = c("Placebo", "Drug", "Placebo", "Drug", "Placebo", "Drug", "Placebo", "Drug", "Placebo", "Drug")
)
```

------------------------------------------------------------------------

## 📈 4. Create Descriptive Stats Table (Continuous Variables)

```{r}
demo %>%
  select(AGE, ARM) %>%
  tbl_summary(by = ARM, statistic = list(all_continuous() ~ "{mean} ± {sd}")) %>%
  add_n() %>%
  add_overall() %>%
  bold_labels()
```

------------------------------------------------------------------------

## 📊 5. Frequency Table for Categorical Variables

```{r}
demo %>%
  select(SEX, ARM) %>%
  tbl_summary(by = ARM) %>%
  add_n() %>%
  add_overall() %>%
  bold_labels()
```

------------------------------------------------------------------------

## 📤 6. Export Summary Table

```{r eval=FALSE}
as_gt(last_tbl()) %>%
  gt::gtsave(filename = "output/summary_table.html")
```

------------------------------------------------------------------------

## 🤖 7. Copilot Prompt Ideas

| Prompt                            | Copilot Suggestion         |
|-----------------------------------|----------------------------|
| `# Summary by treatment arm`      | `tbl_summary(by = ARM)`    |
| `# Format as mean (SD)`           | `"{mean} ± {sd}"`          |
| `# Export table to Word`          | `gt::gtsave()` + `as_gt()` |
| `# Add overall column to summary` | `add_overall()`            |

------------------------------------------------------------------------

## ✅ Summary

| Task                           | R Tool                     |
|--------------------------------|----------------------------|
| Descriptive stats (cont. vars) | `gtsummary::tbl_summary()` |
| Frequencies (cat. vars)        | `gtsummary`                |
| Export table                   | `as_gt()` + `gtsave()`     |
| Copilot formatting support     | Prompt for templates       |
