---
title: "Module 1 Hands-On Exercise"
format: html
editor: visual
---

# 🧪 Getting Started with R --- Hands-On

This document will guide you through your first real R programming task as a clinical programmer.\
You'll create and manipulate a demographics dataset, label it, summarize it, and export it.

------------------------------------------------------------------------

## ✅ Part 1 --- Create a Demographics Dataset

We'll start by creating a simple dataset with 5 subjects.

```{r}
# Create a mock DM dataset
dm <- data.frame(
  USUBJID = c("01-001", "01-002", "01-003", "01-004", "01-005"),
  AGE = c(34, 50, 62, 41, 27),
  SEX = c("M", "F", "M", "F", "F"),
  stringsAsFactors = FALSE
)

# View first few rows
head(dm)
```

------------------------------------------------------------------------

## ✅ Part 2 --- Create Derived Variables

We'll add a few new variables using `mutate()` from the `{dplyr}` package.

```{r}
library(dplyr)

dm <- dm %>%
  mutate(
    AGEGRP = ifelse(AGE < 50, "YOUNG", "OLD"),
    ISFEMALE = ifelse(SEX == "F", "Yes", "No")
  )

dm
```

------------------------------------------------------------------------

## ✅ Part 3 --- Add Variable Labels

Use `{labelled}` to assign labels to the variables, similar to SAS.

```{r}
library(labelled)

var_label(dm$AGE) <- "Age at screening"
var_label(dm$SEX) <- "Biological sex"
var_label(dm$AGEGRP) <- "Age category"
var_label(dm$ISFEMALE) <- "Is subject female?"

# Display labelled dataset
dm
```

------------------------------------------------------------------------

## ✅ Part 4 --- Summarize the Dataset

Let's calculate statistics grouped by variables of interest.

```{r}
# Summary by AGEGRP
dm %>%
  group_by(AGEGRP) %>%
  summarise(
    n = n(),
    mean_age = mean(AGE),
    sd_age = sd(AGE)
  )
```

```{r}
# Count by SEX
dm %>%
  group_by(SEX) %>%
  summarise(n = n())
```

------------------------------------------------------------------------

## ✅ Part 5 --- Export to XPT Format

Save the dataset to a .xpt file, just like in SAS.

```{r}
library(haven)

# Create output folder if it doesn't exist
if (!dir.exists("output")) dir.create("output")

# Export
write_xpt(dm, "output/dm_demo.xpt")
```

------------------------------------------------------------------------

## 💡 Bonus --- Copilot Prompt Practice (Optional)

Try writing the following comment and letting GitHub Copilot suggest code:

``` r
# Flag subjects over 60 as elderly
```

You might get something like:

```{r}
dm <- dm %>%
  mutate(ELDERLY = ifelse(AGE > 60, "Yes", "No"))
```

------------------------------------------------------------------------

## 📊 Bonus --- Plot Distribution by Sex

Create a simple bar plot using `ggplot2`:

```{r}
library(ggplot2)

ggplot(dm, aes(x = SEX)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Sex Distribution", x = "Sex", y = "Count")
```

------------------------------------------------------------------------

## ✅ Summary

In this exercise, you learned how to: - Create and manipulate a data frame - Add derived variables and labels - Summarize your dataset - Export to `.xpt` - Use AI (Copilot) for efficient coding

You're now ready for SDTM programming in Module 2!
