---
title: "Module 1 Theory — R for SAS Programmers: First Contact"
format: html
editor: visual
---

# 🧠 Module 1 Theory --- R for SAS Programmers: First Contact

## 🎯 Learning Objectives

By the end of this section, you will:

-   Understand what R is and how it differs from SAS\
-   Be familiar with the RStudio interface and file types\
-   Learn how to write basic R code (variables, data frames, functions)\
-   Know the key R packages for clinical programming\
-   Understand how to use GitHub Copilot to assist you while coding

------------------------------------------------------------------------

## 🧪 1. What is R?

R is an open-source language designed for data analysis, statistics, and graphics.

Like SAS, it is widely used in clinical research, but:

-   It is object-oriented and script-based\
-   It offers greater flexibility and interactivity\
-   It integrates easily with HTML, PDF, Word, and Shiny dashboards

**SAS vs R Analogy:**

| Feature       | SAS                         | R                            |
|---------------|-----------------------------|------------------------------|
| Interface     | Enterprise Guide / Base SAS | RStudio                      |
| Language Type | Procedural                  | Functional / Object-Oriented |
| Data          | Datasets                    | Data frames / tibbles        |
| Output        | PROC PRINT, ODS             | Console, Plots, Markdown     |
| Extensions    | Macros                      | Packages                     |
| Licensing     | Commercial                  | Free / Open-source           |

------------------------------------------------------------------------

## 🧱 2. RStudio Overview

RStudio is like Enterprise Guide but for R.

### 🖥️ Main Panels:

-   **Script Editor**: Write R scripts (`.R`) or notebooks (`.Rmd`)\
-   **Console**: Execute code\
-   **Environment**: See variables in memory\
-   **Files/Plots/Packages/Help**: Navigate your files and visualize plots

### 📁 File Types:

-   `.R`: regular script (like a SAS `.sas` file)\
-   `.Rmd`: RMarkdown document (mix of code + documentation)\
-   `.Rproj`: RStudio project file to organize files and paths

------------------------------------------------------------------------

## 🧬 3. Writing R Code --- Core Concepts

### ➕ Variables

```{r}
x <- 10          # Assign 10 to x
name <- "Chiara" # Assign string to variable
```

### 🧱 Data Frames (similar to datasets)

```{r}
df <- data.frame(Name = c("John", "Anna"), Age = c(30, 25))
```

### 🔧 Functions

```{r}
mean(c(1, 2, 3))       # Calculates average
sum(c(1, 2, 3))        # Adds values
```

### 💡 Pipe Operator (`%>%`)

Like chaining multiple steps together (e.g., `DATA -> SET -> WHERE`)

```{r}
library(dplyr)

df %>%
  filter(Age > 26) %>%
  select(Name)
```

------------------------------------------------------------------------

## 📦 4. Key R Packages for Clinical Programming

| Package     | Purpose                            |
|-------------|------------------------------------|
| `haven`     | Import/export SAS XPT files        |
| `dplyr`     | Data manipulation (like DATA step) |
| `stringr`   | String handling                    |
| `lubridate` | Date/time manipulation             |
| `admiral`   | Create ADaM datasets               |
| `xportr`    | Label/export SDTM/ADaM to XPT      |
| `ggplot2`   | Plots and visualizations           |
| `flextable` | Export tables to Word/PDF          |

------------------------------------------------------------------------

## 🤖 5. Introduction to GitHub Copilot (Free Plan)

### What is Copilot?

GitHub Copilot is an AI pair programmer that suggests code in real time based on what you type.

### How to Use:

-   Install the Copilot extension in RStudio (via VS Code or browser plugin)\
-   Write a comment or a partial function:

```{r}
# Read a SAS XPT file and convert to tibble
```

-   Press **Tab** to accept Copilot's suggestion.

### Examples:

| Prompt | Copilot Suggestion |
|----|----|
| `# Create a variable for age group` | `mutate(age_group = ifelse(Age < 50, "Y", "O"))` |
| `# Summarize AEs by severity` | `group_by(SEV) %>% summarise(n = n())` |
| `# Convert ISO date to date object` | `mutate(date = as.Date(date, format = "%Y-%m-%d"))` |

> You can always edit or reject suggestions. Copilot is here to help, not to replace your reasoning.

------------------------------------------------------------------------

## 📝 Summary

| Concept           | SAS Equivalent      | R Tool                    |
|-------------------|---------------------|---------------------------|
| Read XPT          | `libname in xpt`    | `haven::read_xpt()`       |
| Data manipulation | `DATA step`         | `dplyr::mutate`, `filter` |
| Export XPT        | `PROC COPY`         | `xportr_write()`          |
| Output tables     | `PROC PRINT/REPORT` | `flextable`, `gt`         |
| AI helper         | Not available       | GitHub Copilot (free)     |
