---
title: "Custom Functions Library"
subtitle: "Reusable R Functions and SAS Macro Translations"
format: 
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: false
  pdf:
    toc: true
    number-sections: true
editor: visual
---

# 🔧 Custom Functions Library

**Reusable R Functions and SAS Macro Translations**

This library provides custom R functions that replicate common SAS macro functionality for clinical programming, along with utility functions for data processing and analysis.

## Overview

This collection includes:

- **SAS Macro Equivalents**: R functions that replicate SAS macro behavior
- **Data Validation Functions**: Quality control and data checking utilities  
- **Date/Time Functions**: Enhanced date manipulation for clinical data
- **String Processing Functions**: Text cleaning and standardization
- **Statistical Functions**: Common calculations for clinical analyses

## Required Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(tibble)
library(stringr)
library(lubridate)
library(purrr)

cat("=== Custom Functions Library ===\n")
cat("Reusable R functions for clinical programming\n\n")
```

## SAS Macro Equivalents

### 1. Frequency Procedure (PROC FREQ equivalent)

```{r}
#| label: proc-freq

#' Calculate frequencies and percentages (like SAS PROC FREQ)
#' @param data Dataset to analyze
#' @param var Variable name to analyze
#' @param by_var Optional grouping variable
#' @param missing Include missing values (default: FALSE)
#' @return Frequency table with counts and percentages

proc_freq <- function(data, var, by_var = NULL, missing = FALSE) {
  
  if (!missing) {
    data <- data %>%
      filter(!is.na(!!sym(var)) & !!sym(var) != "")
  }
  
  if (!is.null(by_var)) {
    # Cross-tabulation
    result <- data %>%
      group_by(!!sym(by_var), !!sym(var)) %>%
      summarise(
        FREQUENCY = n(),
        .groups = "drop"
      ) %>%
      group_by(!!sym(by_var)) %>%
      mutate(
        PERCENT = round(FREQUENCY / sum(FREQUENCY) * 100, 2),
        CUM_FREQ = cumsum(FREQUENCY),
        CUM_PERCENT = round(cumsum(PERCENT), 2)
      ) %>%
      ungroup()
  } else {
    # Simple frequency
    result <- data %>%
      group_by(!!sym(var)) %>%
      summarise(
        FREQUENCY = n(),
        .groups = "drop"
      ) %>%
      mutate(
        PERCENT = round(FREQUENCY / sum(FREQUENCY) * 100, 2),
        CUM_FREQ = cumsum(FREQUENCY),
        CUM_PERCENT = round(cumsum(PERCENT), 2)
      )
  }
  
  return(result)
}

# Example usage
sample_data <- tibble(
  SEX = sample(c("M", "F"), 50, replace = TRUE),
  ARM = sample(c("Placebo", "Treatment"), 50, replace = TRUE),
  AGE_GROUP = sample(c("18-44", "45-64", "65+"), 50, replace = TRUE)
)

print("PROC FREQ example - Sex distribution:")
print(proc_freq(sample_data, "SEX"))

print("\nCross-tabulation - Sex by Treatment:")
print(proc_freq(sample_data, "SEX", "ARM"))
```

### 2. Means Procedure (PROC MEANS equivalent)

```{r}
#| label: proc-means

#' Calculate descriptive statistics (like SAS PROC MEANS)
#' @param data Dataset to analyze
#' @param var Numeric variable to analyze
#' @param by_var Optional grouping variable
#' @param stats Statistics to calculate
#' @return Summary statistics table

proc_means <- function(data, var, by_var = NULL, 
                      stats = c("n", "mean", "std", "min", "median", "max")) {
  
  # Filter out missing values
  data_clean <- data %>%
    filter(!is.na(!!sym(var)))
  
  if (!is.null(by_var)) {
    result <- data_clean %>%
      group_by(!!sym(by_var)) %>%
      summarise(
        N = if ("n" %in% stats) n() else NULL,
        MEAN = if ("mean" %in% stats) round(mean(!!sym(var)), 2) else NULL,
        STD = if ("std" %in% stats) round(sd(!!sym(var)), 2) else NULL,
        MIN = if ("min" %in% stats) min(!!sym(var)) else NULL,
        MEDIAN = if ("median" %in% stats) median(!!sym(var)) else NULL,
        MAX = if ("max" %in% stats) max(!!sym(var)) else NULL,
        .groups = "drop"
      ) %>%
      select_if(~!all(is.null(.)))
  } else {
    result <- data_clean %>%
      summarise(
        N = if ("n" %in% stats) n() else NULL,
        MEAN = if ("mean" %in% stats) round(mean(!!sym(var)), 2) else NULL,
        STD = if ("std" %in% stats) round(sd(!!sym(var)), 2) else NULL,
        MIN = if ("min" %in% stats) min(!!sym(var)) else NULL,
        MEDIAN = if ("median" %in% stats) median(!!sym(var)) else NULL,
        MAX = if ("max" %in% stats) max(!!sym(var)) else NULL
      ) %>%
      select_if(~!all(is.null(.)))
  }
  
  return(result)
}

# Example usage
sample_data$AGE <- sample(18:75, 50, replace = TRUE)

print("PROC MEANS example - Age statistics:")
print(proc_means(sample_data, "AGE"))

print("\nAge statistics by treatment group:")
print(proc_means(sample_data, "AGE", "ARM"))
```

### 3. Data Step Functions

```{r}
#| label: data-step-functions

#' SAS LAG function equivalent
#' @param x Vector to lag
#' @param n Number of periods to lag (default: 1)
#' @return Lagged vector

lag_sas <- function(x, n = 1) {
  c(rep(NA, n), head(x, -n))
}

#' SAS DIF function equivalent  
#' @param x Vector to difference
#' @param n Number of periods for difference (default: 1)
#' @return Differenced vector

dif_sas <- function(x, n = 1) {
  x - lag_sas(x, n)
}

#' SAS FIRST./LAST. variable equivalent
#' @param data Dataset
#' @param by_var Grouping variable
#' @return Dataset with FIRST and LAST indicators

first_last <- function(data, by_var) {
  data %>%
    group_by(!!sym(by_var)) %>%
    mutate(
      FIRST = row_number() == 1,
      LAST = row_number() == n()
    ) %>%
    ungroup()
}

# Example usage
test_data <- tibble(
  USUBJID = rep(c("001", "002", "003"), each = 3),
  VISIT = rep(1:3, 3),
  VALUE = c(10, 12, 15, 8, 11, 9, 20, 18, 22)
)

test_result <- test_data %>%
  mutate(
    LAG_VALUE = lag_sas(VALUE),
    CHANGE = dif_sas(VALUE)
  ) %>%
  first_last("USUBJID")

print("SAS function equivalents:")
print(test_result)
```

## Data Validation Functions

### 1. Missing Data Analysis

```{r}
#| label: missing-analysis

#' Comprehensive missing data analysis
#' @param data Dataset to analyze
#' @param vars Variables to check (default: all)
#' @return Missing data summary

analyze_missing <- function(data, vars = NULL) {
  
  if (is.null(vars)) {
    vars <- names(data)
  }
  
  missing_summary <- data %>%
    select(all_of(vars)) %>%
    summarise_all(~sum(is.na(.) | . == "")) %>%
    pivot_longer(everything(), names_to = "VARIABLE", values_to = "MISSING_COUNT") %>%
    mutate(
      TOTAL_N = nrow(data),
      MISSING_PCT = round(MISSING_COUNT / TOTAL_N * 100, 2),
      COMPLETE_COUNT = TOTAL_N - MISSING_COUNT,
      COMPLETE_PCT = round(COMPLETE_COUNT / TOTAL_N * 100, 2)
    ) %>%
    arrange(desc(MISSING_PCT))
  
  return(missing_summary)
}

# Example usage
sample_data_missing <- sample_data %>%
  mutate(
    AGE = if_else(row_number() %in% sample(1:50, 5), NA_real_, AGE),
    SEX = if_else(row_number() %in% sample(1:50, 2), NA_character_, SEX)
  )

print("Missing data analysis:")
print(analyze_missing(sample_data_missing))
```

### 2. Data Range Validation

```{r}
#| label: range-validation

#' Validate numeric ranges and flag outliers
#' @param data Dataset
#' @param var Variable to validate
#' @param min_val Minimum expected value
#' @param max_val Maximum expected value
#' @param method Outlier detection method
#' @return Validation results

validate_range <- function(data, var, min_val = NULL, max_val = NULL, method = "iqr") {
  
  values <- data[[var]][!is.na(data[[var]])]
  
  results <- list(
    variable = var,
    n_obs = length(values),
    n_missing = sum(is.na(data[[var]])),
    min_observed = min(values, na.rm = TRUE),
    max_observed = max(values, na.rm = TRUE),
    mean_value = round(mean(values, na.rm = TRUE), 2),
    median_value = median(values, na.rm = TRUE)
  )
  
  # Range validation
  if (!is.null(min_val)) {
    below_min <- data %>%
      filter(!is.na(!!sym(var)) & !!sym(var) < min_val)
    results$below_minimum <- nrow(below_min)
    results$below_min_subjects <- below_min
  }
  
  if (!is.null(max_val)) {
    above_max <- data %>%
      filter(!is.na(!!sym(var)) & !!sym(var) > max_val)
    results$above_maximum <- nrow(above_max)
    results$above_max_subjects <- above_max
  }
  
  # Outlier detection
  if (method == "iqr") {
    Q1 <- quantile(values, 0.25)
    Q3 <- quantile(values, 0.75)
    IQR <- Q3 - Q1
    lower_bound <- Q1 - 1.5 * IQR
    upper_bound <- Q3 + 1.5 * IQR
    
    outliers <- data %>%
      filter(!is.na(!!sym(var)) & 
             (!!sym(var) < lower_bound | !!sym(var) > upper_bound))
    
    results$outliers_count <- nrow(outliers)
    results$outliers <- outliers
    results$outlier_bounds <- c(lower_bound, upper_bound)
  }
  
  return(results)
}

# Example usage
age_validation <- validate_range(sample_data_missing, "AGE", min_val = 18, max_val = 80)

print("Age validation results:")
cat("Variable:", age_validation$variable, "\n")
cat("Observations:", age_validation$n_obs, "\n")
cat("Missing:", age_validation$n_missing, "\n")
cat("Range:", age_validation$min_observed, "-", age_validation$max_observed, "\n")
cat("Outliers:", age_validation$outliers_count, "\n")
```

## Date/Time Functions

### 1. Advanced Date Calculations

```{r}
#| label: date-functions

#' Calculate age in years (like SAS YRDIF function)
#' @param birth_date Birth date
#' @param ref_date Reference date (default: today)
#' @param basis Age calculation basis
#' @return Age in years

calculate_age <- function(birth_date, ref_date = Sys.Date(), basis = "actual") {
  
  if (basis == "actual") {
    age <- as.numeric(difftime(ref_date, birth_date, units = "days")) / 365.25
  } else if (basis == "30/360") {
    # Simplified 30/360 calculation
    age <- as.numeric(difftime(ref_date, birth_date, units = "days")) / 360
  } else {
    age <- as.numeric(difftime(ref_date, birth_date, units = "days")) / 365
  }
  
  return(floor(age))
}

#' Study day calculation (like SAS study day)
#' @param event_date Event date
#' @param ref_date Reference date (study start)
#' @return Study day

study_day <- function(event_date, ref_date) {
  diff_days <- as.numeric(event_date - ref_date)
  
  # Study day convention: Day 1 is first day, negative for pre-study days
  study_days <- if_else(diff_days >= 0, diff_days + 1, diff_days)
  
  return(study_days)
}

#' Format dates for CDISC (ISO 8601)
#' @param date_var Date variable
#' @param include_time Include time component
#' @return Formatted date string

format_cdisc_date <- function(date_var, include_time = FALSE) {
  
  if (include_time) {
    # Full ISO 8601 datetime format
    formatted <- format(as.POSIXct(date_var), "%Y-%m-%dT%H:%M:%S")
  } else {
    # Date only
    formatted <- format(as.Date(date_var), "%Y-%m-%d")
  }
  
  return(formatted)
}

# Example usage
date_examples <- tibble(
  USUBJID = paste0("S00", 1:5),
  BIRTH_DATE = as.Date(c("1990-05-15", "1975-12-03", "1988-08-20", "1982-01-10", "1995-09-25")),
  STUDY_START = as.Date("2024-01-15"),
  VISIT_DATE = as.Date(c("2024-01-15", "2024-01-22", "2024-01-10", "2024-01-18", "2024-01-20"))
) %>%
  mutate(
    AGE = calculate_age(BIRTH_DATE, STUDY_START),
    STUDY_DAY = study_day(VISIT_DATE, STUDY_START),
    VISIT_DTC = format_cdisc_date(VISIT_DATE)
  )

print("Date function examples:")
print(date_examples)
```

## String Processing Functions

### 1. Text Cleaning and Standardization

```{r}
#| label: string-functions

#' Clean and standardize text fields
#' @param text_var Text variable to clean
#' @param case_style Case conversion ("upper", "lower", "title", "sentence")
#' @param remove_extra_spaces Remove extra whitespace
#' @return Cleaned text

clean_text <- function(text_var, case_style = "upper", remove_extra_spaces = TRUE) {
  
  # Remove leading/trailing whitespace
  cleaned <- str_trim(text_var)
  
  # Remove extra spaces
  if (remove_extra_spaces) {
    cleaned <- str_squish(cleaned)
  }
  
  # Apply case conversion
  cleaned <- switch(case_style,
                   "upper" = str_to_upper(cleaned),
                   "lower" = str_to_lower(cleaned),
                   "title" = str_to_title(cleaned),
                   "sentence" = str_to_sentence(cleaned),
                   cleaned)
  
  return(cleaned)
}

#' Parse subject ID components
#' @param usubjid Subject ID to parse
#' @param pattern Regex pattern for parsing
#' @return Parsed components

parse_subject_id <- function(usubjid, pattern = "([A-Z0-9]+)-([0-9]+)-([0-9]+)") {
  
  parsed <- str_match(usubjid, pattern)
  
  result <- tibble(
    USUBJID = usubjid,
    STUDY = parsed[, 2],
    SITE = parsed[, 3],
    SUBJECT = parsed[, 4]
  )
  
  return(result)
}

#' Standardize medical terms (simplified MedDRA-like mapping)
#' @param raw_term Raw medical term
#' @return Standardized term

standardize_medical_term <- function(raw_term) {
  
  # Simple term standardization mapping
  term_map <- c(
    "headache" = "Headache",
    "head ache" = "Headache", 
    "nausea" = "Nausea",
    "sick to stomach" = "Nausea",
    "dizzy" = "Dizziness",
    "dizziness" = "Dizziness",
    "tired" = "Fatigue",
    "fatigue" = "Fatigue",
    "sleepy" = "Somnolence"
  )
  
  # Clean and lookup
  clean_term <- str_to_lower(str_trim(raw_term))
  standardized <- term_map[clean_term]
  
  # Return original if no mapping found
  result <- if_else(is.na(standardized), str_to_title(raw_term), standardized)
  
  return(result)
}

# Example usage
text_examples <- tibble(
  RAW_RACE = c("  white  ", "BLACK OR african american", "asian", "White"),
  RAW_AE = c("headache", "sick to stomach", "dizzy", "tired"),
  USUBJID = c("ABC-123-001", "ABC-123-002", "DEF-456-003", "ABC-123-004")
)

text_cleaned <- text_examples %>%
  mutate(
    CLEAN_RACE = clean_text(RAW_RACE, "upper"),
    STANDARD_AE = standardize_medical_term(RAW_AE)
  ) %>%
  bind_cols(parse_subject_id(text_examples$USUBJID))

print("String processing examples:")
print(text_cleaned %>% select(-USUBJID1))  # Remove duplicate column
```

## Statistical Functions

### 1. Clinical Trial Statistics

```{r}
#| label: clinical-stats

#' Calculate confidence intervals for proportions
#' @param x Number of successes
#' @param n Total number of observations
#' @param conf_level Confidence level (default: 0.95)
#' @param method Method for CI calculation
#' @return Confidence interval

prop_ci <- function(x, n, conf_level = 0.95, method = "wilson") {
  
  p <- x / n
  alpha <- 1 - conf_level
  z <- qnorm(1 - alpha/2)
  
  if (method == "wilson") {
    # Wilson score interval
    denominator <- 1 + z^2/n
    center <- (p + z^2/(2*n)) / denominator
    half_width <- z * sqrt((p*(1-p) + z^2/(4*n)) / n) / denominator
    
    lower <- center - half_width
    upper <- center + half_width
    
  } else if (method == "exact") {
    # Exact binomial CI
    if (x == 0) {
      lower <- 0
    } else {
      lower <- qbeta(alpha/2, x, n - x + 1)
    }
    
    if (x == n) {
      upper <- 1
    } else {
      upper <- qbeta(1 - alpha/2, x + 1, n - x)
    }
    
  } else {
    # Normal approximation (Wald)
    se <- sqrt(p * (1 - p) / n)
    lower <- p - z * se
    upper <- p + z * se
  }
  
  return(tibble(
    n = n,
    x = x,
    proportion = round(p, 4),
    lower_ci = round(pmax(0, lower), 4),
    upper_ci = round(pmin(1, upper), 4),
    method = method
  ))
}

#' Calculate treatment effect measures
#' @param data Dataset with treatment and response
#' @param treatment_var Treatment variable
#' @param response_var Response variable (binary)
#' @param treatment_level Active treatment level
#' @param control_level Control treatment level
#' @return Treatment effect measures

treatment_effect <- function(data, treatment_var, response_var, 
                           treatment_level, control_level) {
  
  # Calculate response rates
  summary_stats <- data %>%
    filter(!!sym(treatment_var) %in% c(treatment_level, control_level)) %>%
    group_by(!!sym(treatment_var)) %>%
    summarise(
      n = n(),
      responders = sum(!!sym(response_var) == 1, na.rm = TRUE),
      response_rate = round(responders / n, 4),
      .groups = "drop"
    )
  
  trt_stats <- summary_stats %>% filter(!!sym(treatment_var) == treatment_level)
  ctrl_stats <- summary_stats %>% filter(!!sym(treatment_var) == control_level)
  
  # Calculate effect measures
  risk_difference <- trt_stats$response_rate - ctrl_stats$response_rate
  relative_risk <- trt_stats$response_rate / ctrl_stats$response_rate
  
  if (ctrl_stats$response_rate < 1) {
    odds_ratio <- (trt_stats$response_rate / (1 - trt_stats$response_rate)) / 
                  (ctrl_stats$response_rate / (1 - ctrl_stats$response_rate))
  } else {
    odds_ratio <- NA
  }
  
  return(tibble(
    treatment = treatment_level,
    control = control_level,
    trt_n = trt_stats$n,
    trt_responders = trt_stats$responders,
    trt_rate = trt_stats$response_rate,
    ctrl_n = ctrl_stats$n,
    ctrl_responders = ctrl_stats$responders,
    ctrl_rate = ctrl_stats$response_rate,
    risk_difference = round(risk_difference, 4),
    relative_risk = round(relative_risk, 4),
    odds_ratio = round(odds_ratio, 4)
  ))
}

# Example usage
response_data <- tibble(
  USUBJID = paste0("S", sprintf("%03d", 1:100)),
  ARM = sample(c("Treatment", "Placebo"), 100, replace = TRUE),
  RESPONSE = rbinom(100, 1, prob = ifelse(ARM == "Treatment", 0.6, 0.4))
)

# Calculate confidence intervals for response rates
response_summary <- response_data %>%
  group_by(ARM) %>%
  summarise(
    n = n(),
    responders = sum(RESPONSE),
    .groups = "drop"
  )

ci_results <- map2_dfr(response_summary$responders, response_summary$n, 
                       ~prop_ci(.x, .y, method = "wilson")) %>%
  bind_cols(response_summary %>% select(ARM))

print("Response rate confidence intervals:")
print(ci_results)

# Calculate treatment effect
effect_results <- treatment_effect(response_data, "ARM", "RESPONSE", 
                                 "Treatment", "Placebo")

print("\nTreatment effect measures:")
print(effect_results)
```

## Function Usage Examples

### 1. Complete Data Processing Pipeline

```{r}
#| label: pipeline-example

# Example of using multiple custom functions together
process_clinical_data <- function(raw_data) {
  
  cat("Processing clinical dataset...\n")
  
  # Step 1: Clean and standardize
  processed <- raw_data %>%
    mutate(
      # Clean text fields
      SEX = clean_text(SEX, "upper"),
      RACE = clean_text(RACE, "upper"),
      
      # Calculate derived variables
      AGE = calculate_age(BIRTH_DATE, STUDY_START),
      STUDY_DAY = study_day(VISIT_DATE, STUDY_START),
      
      # Add first/last indicators
      .by_subject = TRUE
    ) %>%
    first_last("USUBJID")
  
  # Step 2: Data validation
  cat("Validation results:\n")
  missing_report <- analyze_missing(processed)
  age_check <- validate_range(processed, "AGE", min_val = 18, max_val = 80)
  
  cat("Missing data summary available\n")
  cat("Age validation complete\n")
  
  # Step 3: Generate summary statistics
  demographics <- proc_means(processed, "AGE", "SEX")
  
  return(list(
    data = processed,
    missing_report = missing_report,
    age_validation = age_check,
    demographics = demographics
  ))
}

# Example dataset
example_raw <- tibble(
  USUBJID = paste0("ABC-123-", sprintf("%03d", 1:20)),
  BIRTH_DATE = sample(seq(as.Date("1950-01-01"), as.Date("1995-12-31"), by = "day"), 20),
  SEX = sample(c("m", "F", "MALE", "female"), 20, replace = TRUE),
  RACE = sample(c("white", "BLACK", "Asian"), 20, replace = TRUE),
  STUDY_START = as.Date("2024-01-15"),
  VISIT_DATE = as.Date("2024-01-15") + sample(0:90, 20, replace = TRUE)
)

# Process the data
results <- process_clinical_data(example_raw)

print("Processed demographics summary:")
print(results$demographics)
```

## Function Documentation

### Best Practices

1. **Function Design**
   - Single responsibility principle
   - Clear parameter names and defaults
   - Comprehensive error handling
   - Consistent return formats

2. **Documentation**
   - Use roxygen2-style comments
   - Include parameter descriptions
   - Provide usage examples
   - Document assumptions and limitations

3. **Testing**
   - Test with edge cases
   - Validate against known results
   - Check for proper error handling
   - Compare with SAS output when applicable

4. **Performance**
   - Use vectorized operations
   - Avoid unnecessary loops
   - Profile for bottlenecks
   - Consider data.table for large datasets

---

*This custom functions library provides a solid foundation for clinical programming in R, with emphasis on SAS compatibility and regulatory requirements.*