---
title: "Data Manipulation Examples"
subtitle: "Practical Data Manipulation Examples with dplyr"
format: 
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: false
  pdf:
    toc: true
    number-sections: true
editor: visual
---

# 📊 Data Manipulation Examples

**Practical Data Manipulation Examples with dplyr**  
Module 2-3: Data Manipulation, Filtering, Joins, and Summaries

This guide demonstrates essential dplyr operations for clinical data analysis, covering core functions and techniques used in data manipulation workflows.

## Overview

Data manipulation is fundamental to clinical programming. This document covers:

- Essential dplyr functions (`select`, `filter`, `mutate`, `summarise`)
- Data reshaping and transformation techniques
- Joining datasets and handling missing data
- Group-wise operations and statistical summaries

## Required Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(tibble)
library(lubridate)
library(stringr)

cat("=== Data Manipulation Examples ===\n")
cat("Essential dplyr operations for data analysis\n\n")
```

## Sample SDTM Datasets

Let's create sample clinical datasets to demonstrate data manipulation techniques.

### Demographics Domain (DM)

```{r}
#| label: dm-data

# Load validated SDTM datasets (in production, these would be from validated sources)
# Demographics Domain (DM)
dm_sdtm <- tibble(
  STUDYID = "ABC-123",
  DOMAIN = "DM",
  USUBJID = paste0("ABC-123-", sprintf("%03d", 1:20)),
  SUBJID = sprintf("%03d", 1:20),
  SITEID = rep(c("001", "002", "003"), length.out = 20),
  BRTHDTC = sample(seq(as.Date("1950-01-01"), as.Date("1990-12-31"), by = "day"), 20),
  SEX = sample(c("M", "F"), 20, replace = TRUE, prob = c(0.6, 0.4)),
  RACE = sample(c("WHITE", "BLACK OR AFRICAN AMERICAN", "ASIAN"), 20, replace = TRUE),
  ETHNIC = sample(c("HISPANIC OR LATINO", "NOT HISPANIC OR LATINO"), 20, replace = TRUE),
  ARMCD = rep(c("PBO", "TRT"), length.out = 20),
  ARM = case_when(
    ARMCD == "PBO" ~ "Placebo",
    ARMCD == "TRT" ~ "Study Drug 10mg"
  ),
  RFSTDTC = as.Date("2024-01-15"),
  RFENDTC = as.Date("2024-04-15")
)

print("Demographics Domain (DM):")
print(dm_sdtm)
```

### Vital Signs Domain (VS)

```{r}
#| label: vs-data

# Vital Signs Domain (VS) - Multiple records per subject
vs_sdtm <- map_dfr(dm_sdtm$USUBJID[1:10], function(subj) {
  visits <- c("BASELINE", "WEEK 2", "WEEK 4", "WEEK 8", "WEEK 12")
  
  tibble(
    STUDYID = "ABC-123",
    DOMAIN = "VS",
    USUBJID = subj,
    VISIT = visits,
    VISITNUM = 1:5,
    VSTESTCD = rep(c("SYSBP", "DIABP", "PULSE"), each = 5),
    VSTEST = rep(c("Systolic Blood Pressure", "Diastolic Blood Pressure", "Pulse Rate"), each = 5),
    VSORRES = c(
      round(rnorm(5, 125, 15)),  # Systolic BP
      round(rnorm(5, 80, 10)),   # Diastolic BP  
      round(rnorm(5, 70, 12))    # Pulse
    ),
    VSORRESU = rep(c("mmHg", "mmHg", "beats/min"), each = 5),
    VSDTC = rep(seq(as.Date("2024-01-15"), by = "2 weeks", length.out = 5), 3)
  )
}) %>%
  arrange(USUBJID, VSTESTCD, VISITNUM)

print("Vital Signs Domain (VS) - First 15 records:")
print(vs_sdtm %>% slice(1:15))
```

### Adverse Events Domain (AE)

```{r}
#| label: ae-data

# Adverse Events Domain (AE)
ae_sdtm <- tibble(
  STUDYID = "ABC-123",
  DOMAIN = "AE", 
  USUBJID = sample(dm_sdtm$USUBJID, 12, replace = TRUE),
  AEDECOD = sample(c("Headache", "Nausea", "Dizziness", "Fatigue", "Insomnia"), 12, replace = TRUE),
  AESEV = sample(c("MILD", "MODERATE", "SEVERE"), 12, replace = TRUE, prob = c(0.6, 0.3, 0.1)),
  AEREL = sample(c("RELATED", "NOT RELATED", "POSSIBLY RELATED"), 12, replace = TRUE),
  AESTDTC = sample(seq(as.Date("2024-01-20"), as.Date("2024-04-10"), by = "day"), 12),
  AEENDTC = sample(seq(as.Date("2024-02-01"), as.Date("2024-04-15"), by = "day"), 12)
) %>%
  arrange(USUBJID, AESTDTC)

print("Adverse Events Domain (AE):")
print(ae_sdtm)
```

## Core dplyr Operations

### 1. Selecting Columns

```{r}
#| label: select-examples

cat("=== SELECT Operations ===\n")

# Select specific columns
dm_basic <- dm_sdtm %>%
  select(USUBJID, SEX, RACE, ARM)

print("Basic demographics (selected columns):")
print(dm_basic)

# Select columns using patterns
dm_dates <- dm_sdtm %>%
  select(USUBJID, ends_with("DTC"))

print("\nDate columns:")
print(dm_dates)

# Select and rename columns
dm_renamed <- dm_sdtm %>%
  select(
    Subject_ID = USUBJID,
    Gender = SEX,
    Treatment = ARM,
    Start_Date = RFSTDTC
  )

print("\nRenamed columns:")
print(head(dm_renamed))
```

### 2. Filtering Rows

```{r}
#| label: filter-examples

cat("\n=== FILTER Operations ===\n")

# Simple filtering
males_only <- dm_sdtm %>%
  filter(SEX == "M")

print("Male subjects only:")
print(males_only %>% select(USUBJID, SEX, ARM))

# Multiple conditions with AND
treatment_males <- dm_sdtm %>%
  filter(SEX == "M" & ARMCD == "TRT")

print("\nTreatment group males:")
print(treatment_males %>% select(USUBJID, SEX, ARM))

# Multiple conditions with OR
minority_subjects <- dm_sdtm %>%
  filter(RACE %in% c("BLACK OR AFRICAN AMERICAN", "ASIAN") | 
         ETHNIC == "HISPANIC OR LATINO")

print("\nMinority subjects:")
print(minority_subjects %>% select(USUBJID, RACE, ETHNIC))

# Filtering with date conditions
recent_births <- dm_sdtm %>%
  filter(BRTHDTC >= as.Date("1980-01-01"))

print("\nSubjects born after 1980:")
print(recent_births %>% select(USUBJID, BRTHDTC) %>% arrange(BRTHDTC))
```

### 3. Creating New Variables

```{r}
#| label: mutate-examples

cat("\n=== MUTATE Operations ===\n")

# Calculate age and create age groups
dm_enhanced <- dm_sdtm %>%
  mutate(
    # Calculate age at study start
    AGE = as.numeric(floor(difftime(RFSTDTC, BRTHDTC, units = "days") / 365.25)),
    
    # Create age groups
    AGEGROUP = case_when(
      AGE < 30 ~ "18-29",
      AGE < 50 ~ "30-49", 
      AGE < 65 ~ "50-64",
      TRUE ~ "65+"
    ),
    
    # Create binary treatment indicator
    TREATMENT_FLAG = if_else(ARMCD == "TRT", 1, 0),
    
    # Format treatment description
    TREATMENT_DESC = str_to_title(ARM),
    
    # Study duration
    STUDY_DAYS = as.numeric(RFENDTC - RFSTDTC)
  )

print("Enhanced demographics with calculated variables:")
print(dm_enhanced %>% 
       select(USUBJID, SEX, AGE, AGEGROUP, TREATMENT_DESC, STUDY_DAYS) %>%
       head(10))
```

### 4. Arranging Data

```{r}
#| label: arrange-examples

cat("\n=== ARRANGE Operations ===\n")

# Sort by single variable
dm_by_age <- dm_enhanced %>%
  arrange(AGE) %>%
  select(USUBJID, SEX, AGE, ARM)

print("Subjects sorted by age (ascending):")
print(head(dm_by_age, 8))

# Sort by multiple variables
dm_multi_sort <- dm_enhanced %>%
  arrange(ARM, desc(AGE)) %>%
  select(USUBJID, ARM, SEX, AGE)

print("\nSorted by treatment, then age (descending):")
print(head(dm_multi_sort, 10))
```

## Group Operations and Summaries

### 1. Basic Grouping and Summarization

```{r}
#| label: group-summarise

cat("\n=== GROUP BY and SUMMARISE Operations ===\n")

# Basic summary by treatment group
dm_summary <- dm_enhanced %>%
  group_by(ARM) %>%
  summarise(
    N = n(),
    Mean_Age = round(mean(AGE), 1),
    Median_Age = median(AGE),
    Min_Age = min(AGE),
    Max_Age = max(AGE),
    SD_Age = round(sd(AGE), 1),
    Female_N = sum(SEX == "F"),
    Female_Pct = round(Female_N / N * 100, 1)
  )

print("Demographics summary by treatment:")
print(dm_summary)

# Multiple grouping variables
site_treatment_summary <- dm_enhanced %>%
  group_by(SITEID, ARM) %>%
  summarise(
    N = n(),
    Mean_Age = round(mean(AGE), 1),
    .groups = "drop"
  )

print("\nSummary by site and treatment:")
print(site_treatment_summary)
```

### 2. Advanced Grouping Operations

```{r}
#| label: advanced-grouping

# Group-wise calculations with mutate
dm_group_calc <- dm_enhanced %>%
  group_by(ARM) %>%
  mutate(
    # Z-score of age within treatment group
    AGE_ZSCORE = round((AGE - mean(AGE)) / sd(AGE), 2),
    
    # Rank within group
    AGE_RANK = rank(AGE),
    
    # Group size
    GROUP_SIZE = n(),
    
    # Age percentile within group
    AGE_PERCENTILE = round(percent_rank(AGE) * 100, 1)
  ) %>%
  ungroup()

print("Group-wise calculations:")
print(dm_group_calc %>% 
       select(USUBJID, ARM, AGE, AGE_ZSCORE, AGE_RANK, AGE_PERCENTILE) %>%
       arrange(ARM, AGE_RANK) %>%
       head(12))
```

## Data Joining Operations

### 1. Inner Joins

```{r}
#| label: joins

cat("\n=== JOIN Operations ===\n")

# Inner join: Only subjects with both demographics and AEs
dm_ae_inner <- dm_sdtm %>%
  inner_join(ae_sdtm, by = "USUBJID") %>%
  select(USUBJID, SEX, ARM, AEDECOD, AESEV, AEREL)

print("Inner join - Subjects with AEs:")
print(dm_ae_inner)

# Left join: All subjects, with AE data where available
dm_ae_left <- dm_sdtm %>%
  left_join(
    ae_sdtm %>% 
      group_by(USUBJID) %>% 
      summarise(
        Total_AEs = n(),
        Severe_AEs = sum(AESEV == "SEVERE"),
        Related_AEs = sum(AEREL == "RELATED"),
        .groups = "drop"
      ),
    by = "USUBJID"
  ) %>%
  mutate(
    Total_AEs = replace_na(Total_AEs, 0),
    Severe_AEs = replace_na(Severe_AEs, 0),
    Related_AEs = replace_na(Related_AEs, 0)
  )

print("\nLeft join - All subjects with AE counts:")
print(dm_ae_left %>% 
       select(USUBJID, ARM, Total_AEs, Severe_AEs, Related_AEs) %>%
       arrange(desc(Total_AEs)))
```

### 2. Complex Joins with Vital Signs

```{r}
#| label: complex-joins

# Join demographics with baseline vital signs
baseline_vitals <- vs_sdtm %>%
  filter(VISIT == "BASELINE") %>%
  select(USUBJID, VSTESTCD, VSORRES) %>%
  pivot_wider(
    names_from = VSTESTCD,
    values_from = VSORRES,
    names_prefix = "BL_"
  )

dm_vitals <- dm_enhanced %>%
  left_join(baseline_vitals, by = "USUBJID") %>%
  filter(!is.na(BL_SYSBP))  # Only subjects with baseline vitals

print("Demographics with baseline vitals:")
print(dm_vitals %>% 
       select(USUBJID, ARM, AGE, BL_SYSBP, BL_DIABP, BL_PULSE) %>%
       head(8))
```

## Advanced Data Manipulation

### 1. Conditional Operations

```{r}
#| label: conditional-ops

cat("\n=== Conditional Operations ===\n")

# Complex conditional logic
dm_risk_assessment <- dm_enhanced %>%
  mutate(
    # Risk categorization based on multiple factors
    RISK_CATEGORY = case_when(
      AGE >= 65 & SEX == "M" ~ "High Risk",
      AGE >= 65 & SEX == "F" ~ "Moderate Risk",
      AGE < 65 & SEX == "M" ~ "Moderate Risk",
      AGE < 65 & SEX == "F" ~ "Low Risk",
      TRUE ~ "Unknown"
    ),
    
    # Eligibility criteria
    ELIGIBLE = case_when(
      AGE < 18 ~ "Too Young",
      AGE > 75 ~ "Too Old", 
      TRUE ~ "Eligible"
    ),
    
    # Priority scoring
    PRIORITY_SCORE = case_when(
      ARMCD == "TRT" & AGE >= 65 ~ 3,
      ARMCD == "TRT" & AGE < 65 ~ 2,
      ARMCD == "PBO" ~ 1,
      TRUE ~ 0
    )
  )

print("Risk assessment and eligibility:")
print(dm_risk_assessment %>% 
       select(USUBJID, AGE, SEX, ARM, RISK_CATEGORY, ELIGIBLE, PRIORITY_SCORE) %>%
       arrange(desc(PRIORITY_SCORE)))
```

### 2. String Manipulation

```{r}
#| label: string-manipulation

# Advanced string operations
dm_strings <- dm_sdtm %>%
  mutate(
    # Extract site region from site ID
    REGION = case_when(
      str_detect(SITEID, "^00[1-2]") ~ "North America",
      str_detect(SITEID, "^003") ~ "Europe",
      TRUE ~ "Other"
    ),
    
    # Format subject ID for display
    DISPLAY_ID = str_replace(USUBJID, "ABC-123-", "Subject "),
    
    # Create composite descriptions
    SUBJECT_DESC = paste(SEX, RACE, "subject in", ARM, "group"),
    
    # Clean and standardize race
    RACE_CLEAN = str_to_title(str_to_lower(RACE))
  )

print("String manipulation examples:")
print(dm_strings %>% 
       select(USUBJID, DISPLAY_ID, REGION, SUBJECT_DESC) %>%
       head(8))
```

### 3. Window Functions

```{r}
#| label: window-functions

# Window functions for rankings and running calculations
vs_analysis <- vs_sdtm %>%
  filter(VSTESTCD == "SYSBP") %>%
  group_by(USUBJID) %>%
  arrange(USUBJID, VISITNUM) %>%
  mutate(
    # Running mean
    RUNNING_MEAN = round(cummean(VSORRES), 1),
    
    # Change from baseline
    CHANGE_BL = VSORRES - first(VSORRES),
    
    # Percent change from baseline  
    PCT_CHANGE_BL = round((VSORRES - first(VSORRES)) / first(VSORRES) * 100, 1),
    
    # Visit rank (within subject)
    VISIT_RANK = row_number(),
    
    # Flag last visit
    LAST_VISIT = if_else(row_number() == n(), "Y", "N")
  ) %>%
  ungroup()

print("Window functions - Systolic BP analysis:")
print(vs_analysis %>% 
       select(USUBJID, VISIT, VSORRES, RUNNING_MEAN, CHANGE_BL, PCT_CHANGE_BL, LAST_VISIT) %>%
       head(15))
```

## Summary Statistics by Group

### 1. Treatment Group Comparisons

```{r}
#| label: treatment-comparison

# Comprehensive treatment group comparison
treatment_comparison <- dm_enhanced %>%
  group_by(ARM) %>%
  summarise(
    # Sample size
    N = n(),
    
    # Age statistics
    Age_Mean = round(mean(AGE), 1),
    Age_SD = round(sd(AGE), 1),
    Age_Median = median(AGE),
    Age_Min = min(AGE),
    Age_Max = max(AGE),
    
    # Gender distribution
    Female_N = sum(SEX == "F"),
    Female_Pct = round(Female_N / N * 100, 1),
    Male_N = sum(SEX == "M"),
    Male_Pct = round(Male_N / N * 100, 1),
    
    # Race distribution
    White_N = sum(RACE == "WHITE"),
    White_Pct = round(White_N / N * 100, 1),
    
    .groups = "drop"
  )

print("Treatment group comparison:")
print(treatment_comparison)
```

### 2. Site-Level Analysis

```{r}
#| label: site-analysis

# Site-level enrollment and demographics
site_analysis <- dm_enhanced %>%
  group_by(SITEID) %>%
  summarise(
    Total_Enrolled = n(),
    Treatment_N = sum(ARMCD == "TRT"),
    Placebo_N = sum(ARMCD == "PBO"),
    Randomization_Ratio = round(Treatment_N / Placebo_N, 2),
    Mean_Age = round(mean(AGE), 1),
    Female_Pct = round(sum(SEX == "F") / n() * 100, 1),
    .groups = "drop"
  ) %>%
  arrange(desc(Total_Enrolled))

print("Site-level analysis:")
print(site_analysis)
```

## Best Practices Summary

### Key dplyr Principles

1. **Pipe Operator (`%>%`)**: Chain operations for readable code
2. **Group Operations**: Use `group_by()` for category-wise calculations  
3. **Joins**: Choose appropriate join type based on analysis needs
4. **Window Functions**: Leverage for rankings and running calculations
5. **Conditional Logic**: Use `case_when()` for complex categorizations

### Performance Tips

- Filter early to reduce data size
- Use `select()` to keep only needed columns
- Leverage vectorized operations over loops
- Group operations are generally efficient
- Consider `data.table` for very large datasets

### Code Organization

- Use meaningful variable names
- Add comments for complex logic
- Break complex operations into steps
- Validate results at each step
- Document assumptions and business rules

---

*This guide demonstrates essential data manipulation techniques for clinical programming using modern R practices with dplyr.*