---
title: "Date & Text Functions"
subtitle: "lubridate and stringr Practical Examples"
format: 
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: false
  pdf:
    toc: true
    number-sections: true
editor: visual
---

# 📅 Date & Text Functions

**lubridate and stringr Practical Examples**

This guide demonstrates practical applications of date/time manipulation with `lubridate` and text processing with `stringr` for clinical programming scenarios.

## Overview

Clinical programming frequently involves:

- **Date Operations**: Study day calculations, age derivations, date validations
- **Text Processing**: Term standardization, ID parsing, data cleaning
- **Format Conversions**: ISO 8601 compliance, controlled terminology
- **Pattern Matching**: Validation rules, data extraction

## Required Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(tibble)
library(tidyr)
library(lubridate)
library(stringr)
library(stringr)
library(readr)

cat("=== Date & Text Functions Guide ===\n")
cat("lubridate and stringr for clinical programming\n\n")
```

## Date/Time Operations with lubridate

### 1. Basic Date Creation and Parsing

```{r}
#| label: date-basics

cat("=== Date Creation and Parsing ===\n")

# Multiple ways to create dates
dates_example <- tibble(
  # From strings with different formats
  date_ymd = ymd("2024-03-15"),
  date_mdy = mdy("03/15/2024"), 
  date_dmy = dmy("15/03/2024"),
  
  # From components
  date_make = make_date(2024, 3, 15),
  
  # Current date/time
  today_date = today(),
  now_datetime = now(),
  
  # From ISO 8601 strings (common in clinical data)
  iso_datetime = ymd_hms("2024-03-15 14:30:25"),
  iso_date_only = ymd("2024-03-15")
)

print("Date creation examples:")
glimpse(dates_example)

# Parsing clinical date formats commonly seen in EDC systems
clinical_dates <- c(
  "2024-01-15",           # ISO standard
  "15-JAN-2024",          # SAS-like format
  "01/15/24",             # US format short year
  "2024-01-15T10:30:00",  # ISO with time
  "15JAN2024:10:30:00"    # SAS datetime format
)

parsed_dates <- tibble(
  original = clinical_dates,
  parsed = c(
    ymd(clinical_dates[1]),
    dmy(clinical_dates[2], locale = "en_US.UTF-8"),
    mdy(clinical_dates[3]),
    ymd_hms(clinical_dates[4]),
    dmy_hms(clinical_dates[5], locale = "en_US.UTF-8")
  )
)

print("\nClinical date parsing:")
print(parsed_dates)
```

### 2. Study Day Calculations

```{r}
#| label: study-days

cat("\n=== Study Day Calculations ===\n")

# Create sample clinical data with study reference dates
study_data <- tibble(
  USUBJID = paste0("STUDY-001-", sprintf("%03d", 1:8)),
  
  # Reference dates
  RFSTDTC = as.Date("2024-01-15"),  # Study start date
  RFENDTC = as.Date("2024-07-15"),  # Study end date
  
  # Event dates
  CONSENT_DATE = RFSTDTC - days(sample(1:14, 8, replace = TRUE)),
  RANDOMIZATION_DATE = RFSTDTC + days(sample(0:7, 8, replace = TRUE)),
  FIRST_DOSE_DATE = RANDOMIZATION_DATE + days(sample(0:2, 8, replace = TRUE)),
  LAST_VISIT_DATE = RFSTDTC + days(sample(80:180, 8, replace = TRUE)),
  AE_START_DATE = FIRST_DOSE_DATE + days(sample(5:60, 8, replace = TRUE))
)

# Calculate study days using different conventions
study_days_calculated <- study_data %>%
  mutate(
    # Standard study day calculation (Day 1 = first day of study)
    CONSENT_STDY = as.numeric(CONSENT_DATE - RFSTDTC),
    RANDOM_STDY = as.numeric(RANDOMIZATION_DATE - RFSTDTC) + 1,
    FIRST_DOSE_STDY = as.numeric(FIRST_DOSE_DATE - RFSTDTC) + 1,
    AE_STDY = as.numeric(AE_START_DATE - RFSTDTC) + 1,
    
    # Relative day calculations
    DAYS_FROM_FIRST_DOSE = as.numeric(AE_START_DATE - FIRST_DOSE_DATE),
    DAYS_TO_STUDY_END = as.numeric(RFENDTC - LAST_VISIT_DATE),
    
    # Treatment duration
    TREATMENT_DURATION = as.numeric(LAST_VISIT_DATE - FIRST_DOSE_DATE),
    
    # Format for CDISC (ISO 8601)
    RFSTDTC_ISO = format(RFSTDTC, "%Y-%m-%d"),
    AE_START_ISO = format(AE_START_DATE, "%Y-%m-%d")
  )

print("Study day calculations:")
print(study_days_calculated %>% 
       select(USUBJID, RFSTDTC, FIRST_DOSE_DATE, AE_START_DATE, 
              FIRST_DOSE_STDY, AE_STDY, DAYS_FROM_FIRST_DOSE))
```

### 3. Age Calculations and Date Arithmetic

```{r}
#| label: age-calculations

cat("\n=== Age Calculations ===\n")

# Demographics with birth dates
demographics <- tibble(
  USUBJID = paste0("STUDY-001-", sprintf("%03d", 1:10)),
  BIRTH_DATE = sample(seq(as.Date("1950-01-01"), as.Date("1995-12-31"), by = "day"), 10),
  ICF_DATE = as.Date("2024-01-10") + days(sample(0:20, 10, replace = TRUE)),
  RFSTDTC = as.Date("2024-01-15")
)

# Multiple age calculation methods
age_calculations <- demographics %>%
  mutate(
    # Age at informed consent
    AGE_AT_ICF = floor(as.numeric(difftime(ICF_DATE, BIRTH_DATE, units = "days")) / 365.25),
    
    # Age at study start (standard)
    AGE = floor(as.numeric(difftime(RFSTDTC, BIRTH_DATE, units = "days")) / 365.25),
    
    # Using lubridate interval for precise calculation
    AGE_PRECISE = floor(time_length(interval(BIRTH_DATE, RFSTDTC), "years")),
    
    # Age in months (for pediatric studies)
    AGE_MONTHS = floor(time_length(interval(BIRTH_DATE, RFSTDTC), "months")),
    
    # Age groups for analysis
    AGEGROUP = case_when(
      AGE < 30 ~ "18-29",
      AGE < 50 ~ "30-49",
      AGE < 65 ~ "50-64", 
      TRUE ~ "65+"
    ),
    
    # Birth year for cohort analysis
    BIRTH_YEAR = year(BIRTH_DATE),
    
    # Days since birth (for very precise calculations)
    DAYS_SINCE_BIRTH = as.numeric(difftime(RFSTDTC, BIRTH_DATE, units = "days"))
  )

print("Age calculations:")
print(age_calculations %>% 
       select(USUBJID, BIRTH_DATE, RFSTDTC, AGE, AGE_PRECISE, AGEGROUP))
```

### 4. Date Validation and Quality Control

```{r}
#| label: date-validation

cat("\n=== Date Validation ===\n")

# Sample data with potential date issues
problematic_dates <- tibble(
  USUBJID = paste0("SUBJ-", 1:8),
  BIRTH_DATE = c(
    as.Date("1975-06-15"),
    as.Date("2010-03-20"),  # Too young
    as.Date("1920-01-01"),  # Too old
    as.Date("1985-12-31"),
    as.Date("1992-02-29"),  # Leap year - valid
    as.Date("1992-08-15"),
    as.Date("1899-01-01"),  # Very old
    as.Date("1988-04-10")
  ),
  STUDY_START = as.Date("2024-01-15"),
  VISIT_DATE = c(
    as.Date("2024-01-20"),
    as.Date("2024-01-10"),  # Before study start
    as.Date("2024-03-15"), 
    as.Date("2023-12-01"),  # Before study start
    as.Date("2024-02-28"),
    as.Date("2024-04-15"),
    as.Date("2025-01-01"),  # Future date
    as.Date("2024-01-18")
  )
)

# Validation checks
date_validation <- problematic_dates %>%
  mutate(
    # Age validation
    AGE = floor(as.numeric(difftime(STUDY_START, BIRTH_DATE, units = "days")) / 365.25),
    AGE_VALID = between(AGE, 18, 85),
    AGE_FLAG = case_when(
      AGE < 18 ~ "TOO_YOUNG",
      AGE > 85 ~ "TOO_OLD",
      TRUE ~ "VALID"
    ),
    
    # Visit date validation
    VISIT_BEFORE_STUDY = VISIT_DATE < STUDY_START,
    VISIT_FUTURE = VISIT_DATE > today(),
    VISIT_FLAG = case_when(
      VISIT_BEFORE_STUDY ~ "BEFORE_STUDY",
      VISIT_FUTURE ~ "FUTURE_DATE",
      TRUE ~ "VALID"
    ),
    
    # Logical consistency checks
    BIRTH_AFTER_VISIT = BIRTH_DATE > VISIT_DATE,
    CONSISTENCY_FLAG = if_else(BIRTH_AFTER_VISIT, "BIRTH_AFTER_VISIT", "VALID")
  )

print("Date validation results:")
print(date_validation %>% 
       select(USUBJID, AGE, AGE_FLAG, VISIT_FLAG, CONSISTENCY_FLAG))

# Summary of validation issues
validation_summary <- date_validation %>%
  summarise(
    total_subjects = n(),
    age_issues = sum(AGE_FLAG != "VALID"),
    visit_issues = sum(VISIT_FLAG != "VALID"),
    consistency_issues = sum(CONSISTENCY_FLAG != "VALID")
  )

print("\nValidation summary:")
print(validation_summary)
```

## Text Processing with stringr

### 1. Basic String Operations

```{r}
#| label: string-basics

cat("\n=== Basic String Operations ===\n")

# Sample clinical text data with common issues
clinical_text <- tibble(
  USUBJID = c("ABC-123-001", "abc-123-002", "ABC-123-003", "ABC 123 004"),
  RAW_RACE = c("  White  ", "BLACK or AFRICAN american", "asian", "White"),
  RAW_AE_TERM = c("Headache", "NAUSEA", "stomach pain", "Dizziness"),
  RAW_MEDICATION = c("Aspirin 81mg", "Tylenol (acetaminophen) 500 MG", "ibuprofen", "Advil"),
  COMMENTS = c("Patient doing well", "SOME MILD SYMPTOMS", "no issues reported", "Follow-up needed")
)

# String cleaning and standardization
text_cleaned <- clinical_text %>%
  mutate(
    # Case conversions
    USUBJID_UPPER = str_to_upper(USUBJID),
    RACE_CLEAN = str_to_upper(str_trim(RAW_RACE)),
    AE_TERM_TITLE = str_to_title(str_to_lower(RAW_AE_TERM)),
    MEDICATION_CLEAN = str_to_title(RAW_MEDICATION),
    COMMENTS_SENTENCE = str_to_sentence(str_to_lower(COMMENTS)),
    
    # Remove extra whitespace
    USUBJID_TRIMMED = str_squish(USUBJID),
    
    # String length for validation
    RACE_LENGTH = str_length(RACE_CLEAN),
    COMMENT_LENGTH = str_length(COMMENTS),
    
    # Check for missing/empty strings
    RACE_MISSING = str_trim(RAW_RACE) == "" | is.na(RAW_RACE),
    
    # Extract numbers from medication strings
    MED_DOSE = str_extract(RAW_MEDICATION, "\\d+"),
    MED_DOSE_NUMERIC = as.numeric(MED_DOSE)
  )

print("String cleaning examples:")
print(text_cleaned %>% 
       select(USUBJID, USUBJID_UPPER, RACE_CLEAN, AE_TERM_TITLE, MED_DOSE_NUMERIC))
```

### 2. Pattern Matching and Extraction

```{r}
#| label: pattern-matching

cat("\n=== Pattern Matching and Extraction ===\n")

# Subject ID parsing and validation
subject_ids <- c(
  "ABC-123-001",
  "DEF-456-002", 
  "GHI-789-003",
  "INVALID-ID",
  "ABC123001",
  "XYZ-999-999"
)

id_analysis <- tibble(
  USUBJID = subject_ids
) %>%
  mutate(
    # Check if ID matches expected pattern
    VALID_FORMAT = str_detect(USUBJID, "^[A-Z]{3}-\\d{3}-\\d{3}$"),
    
    # Extract components using regex groups
    STUDY_CODE = str_extract(USUBJID, "^[A-Z]{3}"),
    SITE_NUMBER = str_extract(USUBJID, "(?<=-)(\\d{3})(?=-)"),
    SUBJECT_NUMBER = str_extract(USUBJID, "\\d{3}$"),
    
    # Alternative extraction using str_match
    ID_COMPONENTS = str_match(USUBJID, "^([A-Z]{3})-(\\d{3})-(\\d{3})$"),
    
    # Count components
    DASH_COUNT = str_count(USUBJID, "-"),
    DIGIT_COUNT = str_count(USUBJID, "\\d"),
    
    # Validation flags
    HAS_STUDY_CODE = !is.na(STUDY_CODE),
    PROPER_LENGTH = str_length(USUBJID) == 11
  )

print("Subject ID analysis:")
print(id_analysis %>% 
       select(USUBJID, VALID_FORMAT, STUDY_CODE, SITE_NUMBER, SUBJECT_NUMBER))
```

### 3. Medical Term Standardization

```{r}
#| label: medical-terms

cat("\n=== Medical Term Standardization ===\n")

# Sample adverse event terms with variations
ae_terms <- tibble(
  RAW_TERM = c(
    "headache",
    "Headache",
    "HEAD ACHE",
    "head pain",
    "nausea",
    "NAUSEA",
    "sick to stomach",
    "feeling sick",
    "dizziness",
    "dizzy",
    "light headed",
    "fatigue",
    "tired",
    "exhausted"
  ),
  SEVERITY = sample(c("MILD", "MODERATE", "SEVERE"), 14, replace = TRUE)
)

# Create standardization mapping
standardize_ae_term <- function(raw_term) {
  # Convert to lowercase for matching
  term_lower <- str_to_lower(str_trim(raw_term))
  
  # Define mapping rules
  standardized <- case_when(
    str_detect(term_lower, "head.*ache|head.*pain") ~ "Headache",
    str_detect(term_lower, "nausea|sick.*stomach|feeling.*sick") ~ "Nausea", 
    str_detect(term_lower, "dizz|light.*head") ~ "Dizziness",
    str_detect(term_lower, "fatigue|tired|exhaust") ~ "Fatigue",
    TRUE ~ str_to_title(raw_term)  # Default to title case
  )
  
  return(standardized)
}

# Apply standardization
ae_standardized <- ae_terms %>%
  mutate(
    STANDARD_TERM = standardize_ae_term(RAW_TERM),
    TERM_LENGTH = str_length(RAW_TERM),
    
    # Create term categories
    TERM_CATEGORY = case_when(
      str_detect(STANDARD_TERM, "Headache") ~ "Neurological",
      str_detect(STANDARD_TERM, "Nausea") ~ "Gastrointestinal",
      str_detect(STANDARD_TERM, "Dizziness") ~ "Neurological", 
      str_detect(STANDARD_TERM, "Fatigue") ~ "General",
      TRUE ~ "Other"
    )
  )

print("AE term standardization:")
print(ae_standardized)

# Summary of standardization results
standardization_summary <- ae_standardized %>%
  count(STANDARD_TERM, TERM_CATEGORY, name = "FREQUENCY") %>%
  arrange(desc(FREQUENCY))

print("\nStandardization summary:")
print(standardization_summary)
```

### 4. Data Validation with Regex

```{r}
#| label: data-validation-regex

cat("\n=== Data Validation with Regex ===\n")

# Sample clinical data needing validation
validation_data <- tibble(
  USUBJID = c("ABC-123-001", "DEF-456-002", "invalid", "GHI-789-003"),
  EMAIL = c("john.doe@email.com", "invalid-email", "jane@hospital.org", "test@test"),
  PHONE = c("555-123-4567", "5551234567", "555.123.4567", "invalid"),
  MRN = c("MRN123456", "123456", "MRN-789012", "INVALID123"),
  DOSE_TEXT = c("5 mg", "10mg", "2.5 mg BID", "unknown")
)

# Validation patterns
validation_results <- validation_data %>%
  mutate(
    # Subject ID validation
    USUBJID_VALID = str_detect(USUBJID, "^[A-Z]{3}-\\d{3}-\\d{3}$"),
    
    # Email validation (simplified)
    EMAIL_VALID = str_detect(EMAIL, "^[\\w\\.-]+@[\\w\\.-]+\\.[a-zA-Z]{2,}$"),
    
    # Phone number validation (US format)
    PHONE_VALID = str_detect(PHONE, "^\\d{3}[-.\\s]?\\d{3}[-.\\s]?\\d{4}$"),
    
    # Medical record number validation
    MRN_VALID = str_detect(MRN, "^MRN\\d{6}$"),
    
    # Dose extraction and validation
    DOSE_NUMBER = as.numeric(str_extract(DOSE_TEXT, "\\d+\\.?\\d*")),
    DOSE_UNIT = str_extract(DOSE_TEXT, "mg|g|mL"),
    DOSE_VALID = !is.na(DOSE_NUMBER) & !is.na(DOSE_UNIT),
    
    # Overall validation score
    VALIDATION_SCORE = USUBJID_VALID + EMAIL_VALID + PHONE_VALID + MRN_VALID + DOSE_VALID,
    
    # Validation status
    VALIDATION_STATUS = case_when(
      VALIDATION_SCORE == 5 ~ "PASS",
      VALIDATION_SCORE >= 3 ~ "WARNING", 
      TRUE ~ "FAIL"
    )
  )

print("Data validation results:")
print(validation_results %>% 
       select(USUBJID, USUBJID_VALID, EMAIL_VALID, PHONE_VALID, 
              DOSE_NUMBER, DOSE_UNIT, VALIDATION_STATUS))
```

## Advanced Date/Time Operations

### 1. Time Zone Handling

```{r}
#| label: timezone-handling

cat("\n=== Time Zone Handling ===\n")

# Clinical trials often span multiple time zones
multi_site_data <- tibble(
  SITE_ID = c("001", "002", "003", "004"),
  SITE_NAME = c("New York", "London", "Tokyo", "Sydney"),
  TIMEZONE = c("America/New_York", "Europe/London", "Asia/Tokyo", "Australia/Sydney"),
  LOCAL_TIME = c("2024-03-15 14:30:00", "2024-03-15 19:30:00", 
                 "2024-03-16 03:30:00", "2024-03-16 05:30:00")
)

# Convert to standard UTC time
utc_times <- multi_site_data %>%
  mutate(
    # Parse local time with timezone (simplified approach)
    LOCAL_DATETIME = ymd_hms(LOCAL_TIME),
    
    # Format for CDISC submission (ISO 8601)
    UTC_ISO = format(LOCAL_DATETIME, "%Y-%m-%dT%H:%M:%S"),
    
    # Study day based on date
    STUDY_START = ymd("2024-03-15"),
    STUDY_DAY = as.numeric(date(LOCAL_DATETIME) - STUDY_START) + 1
  )

print("Multi-site time zone handling:")
print(utc_times %>% select(SITE_NAME, LOCAL_TIME, UTC_ISO, STUDY_DAY))
```

### 2. Date Sequences and Intervals

```{r}
#| label: date-sequences

cat("\n=== Date Sequences and Intervals ===\n")

# Create visit schedules
study_schedule <- tibble(
  USUBJID = "ABC-123-001",
  STUDY_START = as.Date("2024-01-15")
) %>%
  mutate(
    # Create visit sequence
    VISIT_DATES = list(STUDY_START + days(c(0, 14, 28, 56, 84, 112, 168))),
    VISIT_NAMES = list(c("Baseline", "Week 2", "Week 4", "Week 8", 
                        "Week 12", "Week 16", "Week 24"))
  ) %>%
  unnest(c(VISIT_DATES, VISIT_NAMES)) %>%
  mutate(
    VISIT_NUMBER = row_number(),
    STUDY_DAY = as.numeric(VISIT_DATES - STUDY_START) + 1,
    
    # Calculate visit windows (±3 days)
    WINDOW_START = VISIT_DATES - days(3),
    WINDOW_END = VISIT_DATES + days(3),
    
    # Day of week for scheduling
    VISIT_DOW = wday(VISIT_DATES, label = TRUE),
    
    # Month and quarter for analysis
    VISIT_MONTH = month(VISIT_DATES, label = TRUE),
    VISIT_QUARTER = quarter(VISIT_DATES)
  )

print("Study visit schedule:")
print(study_schedule)

# Calculate intervals between visits
visit_intervals <- study_schedule %>%
  mutate(
    # Days since previous visit
    DAYS_SINCE_PREV = as.numeric(VISIT_DATES - lag(VISIT_DATES)),
    
    # Weeks since baseline
    WEEKS_SINCE_BL = as.numeric(VISIT_DATES - first(VISIT_DATES)) / 7,
    
    # Duration from baseline
    DURATION_FROM_BL = interval(first(VISIT_DATES), VISIT_DATES),
    DURATION_WEEKS = time_length(DURATION_FROM_BL, "weeks")
  )

print("\nVisit intervals:")
print(visit_intervals %>% 
       select(VISIT_NAMES, VISIT_DATES, DAYS_SINCE_PREV, WEEKS_SINCE_BL))
```

## Advanced String Operations

### 1. Complex Text Parsing

```{r}
#| label: complex-parsing

cat("\n=== Complex Text Parsing ===\n")

# Sample medication text with complex formatting
medication_text <- tibble(
  MED_TEXT = c(
    "Aspirin 81 mg once daily",
    "Metformin 500mg twice daily (BID)", 
    "Lisinopril 10 mg PO QD #30",
    "Atorvastatin 20mg by mouth at bedtime",
    "Ibuprofen 400 mg every 6 hours as needed",
    "Insulin glargine 15 units subcutaneous at bedtime"
  )
)

# Extract medication components
med_parsed <- medication_text %>%
  mutate(
    # Extract medication name (first word(s) before dose)
    MED_NAME = str_extract(MED_TEXT, "^[A-Za-z\\s]+(?=\\s\\d)"),
    MED_NAME_CLEAN = str_trim(MED_NAME),
    
    # Extract dose and unit
    DOSE_WITH_UNIT = str_extract(MED_TEXT, "\\d+\\.?\\d*\\s*mg|\\d+\\.?\\d*\\s*units"),
    DOSE_NUMBER = as.numeric(str_extract(DOSE_WITH_UNIT, "\\d+\\.?\\d*")),
    DOSE_UNIT = str_extract(DOSE_WITH_UNIT, "mg|units"),
    
    # Extract frequency information
    FREQUENCY = case_when(
      str_detect(MED_TEXT, "once daily|QD") ~ "Once daily",
      str_detect(MED_TEXT, "twice daily|BID") ~ "Twice daily",
      str_detect(MED_TEXT, "every 6 hours") ~ "Every 6 hours",
      str_detect(MED_TEXT, "at bedtime") ~ "At bedtime",
      TRUE ~ "As directed"
    ),
    
    # Extract route of administration
    ROUTE = case_when(
      str_detect(MED_TEXT, "PO|by mouth") ~ "Oral",
      str_detect(MED_TEXT, "subcutaneous") ~ "Subcutaneous",
      TRUE ~ "Oral"  # Default assumption
    ),
    
    # Identify PRN (as needed) medications
    PRN_FLAG = str_detect(MED_TEXT, "as needed|PRN"),
    
    # Create standardized medication string
    MED_STANDARD = paste(MED_NAME_CLEAN, DOSE_NUMBER, DOSE_UNIT, FREQUENCY, sep = " ")
  )

print("Medication parsing:")
print(med_parsed %>% 
       select(MED_NAME_CLEAN, DOSE_NUMBER, DOSE_UNIT, FREQUENCY, ROUTE))
```

### 2. Text Quality Assessment

```{r}
#| label: text-quality

cat("\n=== Text Quality Assessment ===\n")

# Sample clinical comments with quality issues
clinical_comments <- tibble(
  COMMENT_ID = 1:10,
  RAW_COMMENT = c(
    "Patient doing well",
    "PATIENT EXPERIENCED MILD HEADACHE",
    "no adverse events reported",
    "",  # Empty comment
    "Patient c/o nausea & dizziness",
    "follow-up needed ASAP",
    "Pt. stable, no issues",
    "VITAL SIGNS WNL",
    "Patient reports feeling much better today",
    "   "  # Whitespace only
  )
)

# Quality assessment
comment_quality <- clinical_comments %>%
  mutate(
    # Basic quality metrics
    COMMENT_LENGTH = str_length(RAW_COMMENT),
    WORD_COUNT = str_count(RAW_COMMENT, "\\w+"),
    
    # Content checks
    IS_EMPTY = str_trim(RAW_COMMENT) == "" | is.na(RAW_COMMENT),
    IS_ALL_CAPS = str_detect(RAW_COMMENT, "^[A-Z\\s\\W]*$") & COMMENT_LENGTH > 0,
    IS_ALL_LOWER = str_detect(RAW_COMMENT, "^[a-z\\s\\W]*$") & COMMENT_LENGTH > 0,
    
    # Medical abbreviation detection
    HAS_ABBREVIATIONS = str_detect(RAW_COMMENT, "\\bc/o\\b|\\bpt\\.?\\b|\\bWNL\\b|\\bASAP\\b"),
    
    # Quality score (0-5)
    QUALITY_SCORE = case_when(
      IS_EMPTY ~ 0,
      COMMENT_LENGTH < 5 ~ 1,
      IS_ALL_CAPS | IS_ALL_LOWER ~ 2,
      HAS_ABBREVIATIONS ~ 3,
      WORD_COUNT < 3 ~ 3,
      TRUE ~ 5
    ),
    
    # Quality category
    QUALITY_CATEGORY = case_when(
      QUALITY_SCORE == 0 ~ "Missing",
      QUALITY_SCORE <= 2 ~ "Poor",
      QUALITY_SCORE == 3 ~ "Fair",
      QUALITY_SCORE >= 4 ~ "Good"
    ),
    
    # Cleaned version
    COMMENT_CLEAN = str_to_sentence(str_squish(RAW_COMMENT))
  )

print("Comment quality assessment:")
print(comment_quality %>% 
       select(RAW_COMMENT, COMMENT_LENGTH, WORD_COUNT, QUALITY_CATEGORY))

# Quality summary
quality_summary <- comment_quality %>%
  count(QUALITY_CATEGORY, name = "COUNT") %>%
  mutate(PERCENTAGE = round(COUNT / sum(COUNT) * 100, 1))

print("\nQuality distribution:")
print(quality_summary)
```

## Integration Examples

### 1. Complete Data Processing Pipeline

```{r}
#| label: integration-pipeline

cat("\n=== Complete Data Processing Pipeline ===\n")

# Simulate raw clinical data with common issues
raw_clinical_data <- tibble(
  subject_id = c("abc-123-001", "ABC-123-002", "abc 123 003", "ABC-123-004"),
  birth_date = c("1985-06-15", "15-JUN-1970", "1990/03/20", "1988-12-01"),
  consent_date = c("2024-01-10 09:30", "2024-01-12 14:15", "2024-01-08 11:00", "2024-01-15 10:45"),
  adverse_event = c("  headache  ", "NAUSEA", "stomach pain", ""),
  medication = c("Aspirin 81mg daily", "Tylenol 500 MG PRN", "Ibuprofen", "Metformin 500mg BID"),
  comments = c("patient doing well", "MILD SYMPTOMS ONLY", "no issues", "follow-up needed")
)

# Comprehensive cleaning pipeline
process_clinical_data <- function(data) {
  processed <- data %>%
    mutate(
      # Standardize subject IDs
      USUBJID = str_to_upper(str_replace_all(subject_id, "\\s+", "-")),
      
      # Parse and standardize dates
      BIRTH_DATE = case_when(
        str_detect(birth_date, "^\\d{4}-\\d{2}-\\d{2}$") ~ ymd(birth_date),
        str_detect(birth_date, "^\\d{2}-[A-Z]{3}-\\d{4}$") ~ dmy(birth_date),
        str_detect(birth_date, "^\\d{4}/\\d{2}/\\d{2}$") ~ ymd(birth_date),
        TRUE ~ as.Date(NA)
      ),
      
      CONSENT_DATETIME = ymd_hm(consent_date),
      CONSENT_DATE = date(CONSENT_DATETIME),
      
      # Calculate age at consent
      AGE_AT_CONSENT = floor(time_length(interval(BIRTH_DATE, CONSENT_DATE), "years")),
      
      # Standardize adverse events
      AE_TERM_CLEAN = case_when(
        str_trim(adverse_event) == "" ~ NA_character_,
        str_detect(str_to_lower(adverse_event), "head") ~ "Headache",
        str_detect(str_to_lower(adverse_event), "nausea|stomach") ~ "Nausea", 
        TRUE ~ str_to_title(str_trim(adverse_event))
      ),
      
      # Parse medications
      MED_NAME = str_extract(medication, "^[A-Za-z\\s]+"),
      MED_DOSE = as.numeric(str_extract(medication, "\\d+")),
      MED_UNIT = str_extract(medication, "mg|g"),
      
      # Clean comments
      COMMENTS_CLEAN = str_to_sentence(str_squish(comments)),
      
      # Data quality flags
      BIRTH_DATE_VALID = !is.na(BIRTH_DATE),
      AGE_REASONABLE = between(AGE_AT_CONSENT, 18, 90),
      HAS_AE = !is.na(AE_TERM_CLEAN),
      
      # Study day calculation
      STUDY_START = as.Date("2024-01-15"),
      CONSENT_STUDY_DAY = as.numeric(CONSENT_DATE - STUDY_START)
    ) %>%
    
    # Select final variables
    select(USUBJID, BIRTH_DATE, CONSENT_DATE, AGE_AT_CONSENT, AE_TERM_CLEAN,
           MED_NAME, MED_DOSE, MED_UNIT, COMMENTS_CLEAN, 
           BIRTH_DATE_VALID, AGE_REASONABLE, HAS_AE, CONSENT_STUDY_DAY)
  
  return(processed)
}

# Process the data
processed_data <- process_clinical_data(raw_clinical_data)

print("Processed clinical data:")
print(processed_data)

# Data quality summary
quality_check <- processed_data %>%
  summarise(
    total_subjects = n(),
    valid_birth_dates = sum(BIRTH_DATE_VALID),
    reasonable_ages = sum(AGE_REASONABLE),
    subjects_with_ae = sum(HAS_AE),
    data_quality_score = round(mean(BIRTH_DATE_VALID + AGE_REASONABLE) * 100, 1)
  )

print("\nData quality summary:")
print(quality_check)
```

## Best Practices Summary

### Date/Time Best Practices

1. **Standardization**
   - Always use ISO 8601 format for storage (`YYYY-MM-DD`)
   - Handle time zones explicitly in multi-site studies
   - Use lubridate functions for reliable parsing

2. **Validation**
   - Check date ranges for clinical plausibility
   - Validate chronological order of events
   - Handle leap years and edge cases properly

3. **Calculations**
   - Use precise interval calculations for age
   - Follow study protocol for study day conventions
   - Document assumptions about missing dates

### Text Processing Best Practices

1. **Cleaning**
   - Remove extra whitespace systematically
   - Standardize case appropriately for context
   - Handle missing/empty strings explicitly

2. **Validation**
   - Use regex patterns for format validation
   - Implement controlled terminology mapping
   - Check for common data entry errors

3. **Documentation**
   - Document regex patterns clearly
   - Maintain mappings for term standardization
   - Track data transformation steps

---

*This guide provides practical examples of date/time and text processing techniques essential for clinical programming in R.*