---
title: "QC Validation Toolkit"
subtitle: "AI-Assisted Quality Control for Clinical Programming"
format: 
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: false
  pdf:
    toc: true
    number-sections: true
editor: visual
---

# 🔍 QC Validation Toolkit

**AI-Assisted Quality Control for Clinical Programming**  
GitHub Copilot Integration for Automated QC Procedures

This toolkit demonstrates comprehensive QC procedures using R with GitHub Copilot assistance for clinical programming validation.

## Overview

Quality Control (QC) is a critical component of clinical programming that ensures data integrity, regulatory compliance, and statistical accuracy. This toolkit provides:

- Automated data validation procedures
- AI-assisted error detection using GitHub Copilot
- Standardized QC report generation
- Cross-domain validation checks

## Required Libraries

```{r}
#| label: setup
#| message: false
#| warning: false

library(dplyr)
library(tibble)
library(stringr)
library(lubridate)
library(tidyr)
library(purrr)

cat("=== AI-Assisted Clinical Data Quality Control ===\n")
cat("Comprehensive QC procedures with GitHub Copilot integration\n\n")
```

## Mock Clinical Data with Intentional Issues

We'll create test data with various quality issues to demonstrate our QC procedures.

```{r}
#| label: mock-data

set.seed(2024)

# Create ADSL with various data quality issues
adsl_qc <- tibble(
  STUDYID = "ABC-123",
  USUBJID = c(
    paste0("ABC-123-", sprintf("%03d", 1:18)),
    "ABC-123-019",  # Duplicate will be added
    "ABC-123-019"   # Duplicate subject
  ),
  SUBJID = c(sprintf("%03d", 1:18), "019", "019"),
  AGE = c(sample(18:75, 18, replace = TRUE), 65, NA),  # Missing age
  SEX = c(sample(c("M", "F"), 18, replace = TRUE), "X", "F"),  # Invalid sex code
  RACE = c(sample(c("WHITE", "BLACK OR AFRICAN AMERICAN", "ASIAN"), 18, replace = TRUE), 
           "white", "WHITE"),  # Inconsistent case
  TRT01P = c(sample(c("Placebo", "Study Drug 5mg", "Study Drug 10mg"), 18, replace = TRUE),
             "PLACEBO", "Study Drug 5mg"),  # Inconsistent treatment name
  TRT01PN = c(sample(c(0, 5, 10), 18, replace = TRUE), 0, 5),
  TRTSDT = c(
    rep(as.Date("2024-01-15"), 18),
    as.Date("2023-12-01"),  # Treatment start before study start
    as.Date("2024-01-15")
  ),
  TRTEDT = c(
    rep(as.Date("2024-03-15"), 18),
    as.Date("2024-03-15"),
    NA  # Missing treatment end date
  ),
  SAFFL = c(rep("Y", 19), ""),  # Missing safety flag
  ITTFL = c(rep("Y", 18), "N", "Y"),
  COMPLFL = c(rep("Y", 17), "N", "Y", "Y"),
  DTHFL = c(rep("N", 19), "Y"),
  DTHDTC = c(rep("", 19), "2024-02-28"),
  DTHDT = c(rep(as.Date(NA), 19), as.Date("2024-02-28"))
)

# Display the problematic data
print("ADSL Data with Quality Issues:")
print(adsl_qc)
```

## QC Check Functions

### 1. Duplicate Subject Check

```{r}
#| label: duplicate-check

check_duplicates <- function(data, id_var = "USUBJID") {
  cat("=== Duplicate Subject Check ===\n")
  
  duplicates <- data %>%
    group_by(!!sym(id_var)) %>%
    filter(n() > 1) %>%
    ungroup()
  
  if (nrow(duplicates) > 0) {
    cat("❌ ISSUE FOUND: Duplicate subjects detected\n")
    cat("Number of duplicate records:", nrow(duplicates), "\n")
    cat("Duplicate subjects:\n")
    print(duplicates %>% select(!!sym(id_var), SUBJID, AGE, SEX))
    
    return(list(
      status = "FAIL",
      count = nrow(duplicates),
      details = duplicates
    ))
  } else {
    cat("✅ PASS: No duplicate subjects found\n")
    return(list(status = "PASS", count = 0, details = NULL))
  }
}

# Run duplicate check
dup_result <- check_duplicates(adsl_qc)
```

### 2. Missing Critical Variables Check

```{r}
#| label: missing-check

check_missing_critical <- function(data, critical_vars = c("USUBJID", "AGE", "SEX", "TRTSDT")) {
  cat("\n=== Missing Critical Variables Check ===\n")
  
  issues <- list()
  
  for (var in critical_vars) {
    if (var %in% names(data)) {
      missing_count <- sum(is.na(data[[var]]) | data[[var]] == "")
      missing_pct <- round(missing_count / nrow(data) * 100, 2)
      
      if (missing_count > 0) {
        cat("❌", var, ":", missing_count, "missing values (", missing_pct, "%)\n")
        
        # Show which subjects have missing values
        missing_subjects <- data %>%
          filter(is.na(!!sym(var)) | !!sym(var) == "") %>%
          select(USUBJID, !!sym(var))
        
        issues[[var]] <- list(
          count = missing_count,
          percentage = missing_pct,
          subjects = missing_subjects
        )
      } else {
        cat("✅", var, ": No missing values\n")
      }
    } else {
      cat("⚠️", var, ": Variable not found in dataset\n")
    }
  }
  
  return(issues)
}

# Run missing values check
missing_result <- check_missing_critical(adsl_qc)
```

### 3. Controlled Terminology Validation

```{r}
#| label: terminology-check

check_controlled_terminology <- function(data) {
  cat("\n=== Controlled Terminology Validation ===\n")
  
  # Define expected controlled terminology
  valid_sex <- c("M", "F")
  valid_race <- c("WHITE", "BLACK OR AFRICAN AMERICAN", "ASIAN", 
                  "AMERICAN INDIAN OR ALASKA NATIVE", "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER")
  valid_flags <- c("Y", "N", "")
  
  issues <- list()
  
  # Check SEX
  invalid_sex <- data %>%
    filter(!SEX %in% valid_sex) %>%
    select(USUBJID, SEX)
  
  if (nrow(invalid_sex) > 0) {
    cat("❌ SEX: Invalid values found\n")
    print(invalid_sex)
    issues$SEX <- invalid_sex
  } else {
    cat("✅ SEX: All values valid\n")
  }
  
  # Check RACE (case sensitivity)
  invalid_race <- data %>%
    filter(!RACE %in% valid_race) %>%
    select(USUBJID, RACE)
  
  if (nrow(invalid_race) > 0) {
    cat("❌ RACE: Invalid/inconsistent values found\n")
    print(invalid_race)
    issues$RACE <- invalid_race
  } else {
    cat("✅ RACE: All values valid\n")
  }
  
  # Check flags
  flag_vars <- c("SAFFL", "ITTFL", "COMPLFL", "DTHFL")
  for (flag_var in flag_vars) {
    if (flag_var %in% names(data)) {
      invalid_flags <- data %>%
        filter(!is.na(!!sym(flag_var)) & !!!sym(flag_var) %in% valid_flags) %>%
        select(USUBJID, !!sym(flag_var))
      
      if (nrow(invalid_flags) > 0) {
        cat("❌", flag_var, ": Invalid flag values found\n")
        print(invalid_flags)
        issues[[flag_var]] <- invalid_flags
      } else {
        cat("✅", flag_var, ": All flag values valid\n")
      }
    }
  }
  
  return(issues)
}

# Run controlled terminology check
ct_result <- check_controlled_terminology(adsl_qc)
```

### 4. Date Logic Validation

```{r}
#| label: date-logic-check

check_date_logic <- function(data) {
  cat("\n=== Date Logic Validation ===\n")
  
  issues <- list()
  study_start <- as.Date("2024-01-01")  # Define study start date
  
  # Check treatment start vs study start
  early_treatment <- data %>%
    filter(!is.na(TRTSDT) & TRTSDT < study_start) %>%
    select(USUBJID, TRTSDT)
  
  if (nrow(early_treatment) > 0) {
    cat("❌ Treatment start before study start:\n")
    print(early_treatment)
    issues$early_treatment <- early_treatment
  } else {
    cat("✅ All treatment start dates valid\n")
  }
  
  # Check treatment end vs treatment start
  invalid_treatment_duration <- data %>%
    filter(!is.na(TRTSDT) & !is.na(TRTEDT) & TRTEDT < TRTSDT) %>%
    select(USUBJID, TRTSDT, TRTEDT)
  
  if (nrow(invalid_treatment_duration) > 0) {
    cat("❌ Treatment end before treatment start:\n")
    print(invalid_treatment_duration)
    issues$invalid_duration <- invalid_treatment_duration
  } else {
    cat("✅ All treatment durations valid\n")
  }
  
  # Check death date logic
  death_logic_issues <- data %>%
    filter(DTHFL == "Y" & (is.na(DTHDT) | DTHDT == "")) %>%
    select(USUBJID, DTHFL, DTHDT)
  
  if (nrow(death_logic_issues) > 0) {
    cat("❌ Death flag Y but missing death date:\n")
    print(death_logic_issues)
    issues$death_logic <- death_logic_issues
  } else {
    cat("✅ Death date logic consistent\n")
  }
  
  return(issues)
}

# Run date logic check
date_result <- check_date_logic(adsl_qc)
```

### 5. Age Validation

```{r}
#| label: age-validation

check_age_validation <- function(data) {
  cat("\n=== Age Validation ===\n")
  
  issues <- list()
  
  # Check age range
  invalid_age <- data %>%
    filter(!is.na(AGE) & (AGE < 18 | AGE > 100)) %>%
    select(USUBJID, AGE)
  
  if (nrow(invalid_age) > 0) {
    cat("❌ Age outside expected range (18-100):\n")
    print(invalid_age)
    issues$age_range <- invalid_age
  } else {
    cat("✅ All ages within expected range\n")
  }
  
  # Age distribution summary
  cat("\nAge Distribution Summary:\n")
  age_summary <- data %>%
    filter(!is.na(AGE)) %>%
    summarise(
      n = n(),
      mean_age = round(mean(AGE), 1),
      median_age = median(AGE),
      min_age = min(AGE),
      max_age = max(AGE),
      sd_age = round(sd(AGE), 1)
    )
  
  print(age_summary)
  
  return(issues)
}

# Run age validation
age_result <- check_age_validation(adsl_qc)
```

## Cross-Domain Validation

```{r}
#| label: cross-domain

# Mock additional domain for cross-validation
adae_qc <- tibble(
  USUBJID = c(
    sample(adsl_qc$USUBJID[1:15], 10),  # Some subjects from ADSL
    "ABC-123-999"  # Subject not in ADSL
  ),
  AEDECOD = sample(c("Headache", "Nausea", "Dizziness", "Fatigue"), 11, replace = TRUE),
  AESEV = sample(c("MILD", "MODERATE", "SEVERE"), 11, replace = TRUE),
  AEREL = sample(c("RELATED", "NOT RELATED", "POSSIBLY RELATED"), 11, replace = TRUE)
)

check_cross_domain <- function(adsl_data, ae_data) {
  cat("\n=== Cross-Domain Validation ===\n")
  
  issues <- list()
  
  # Check for AE subjects not in ADSL
  ae_subjects <- unique(ae_data$USUBJID)
  adsl_subjects <- unique(adsl_data$USUBJID)
  
  orphan_ae <- setdiff(ae_subjects, adsl_subjects)
  
  if (length(orphan_ae) > 0) {
    cat("❌ AE subjects not found in ADSL:\n")
    cat(paste(orphan_ae, collapse = ", "), "\n")
    issues$orphan_ae <- orphan_ae
  } else {
    cat("✅ All AE subjects found in ADSL\n")
  }
  
  # Check for ADSL subjects with safety flag Y but no AEs
  safety_subjects <- adsl_data %>%
    filter(SAFFL == "Y") %>%
    pull(USUBJID)
  
  subjects_no_ae <- setdiff(safety_subjects, ae_subjects)
  
  cat("Safety population subjects without AEs:", length(subjects_no_ae), "\n")
  if (length(subjects_no_ae) > 5) {  # Only show first 5
    cat("Examples:", paste(head(subjects_no_ae, 5), collapse = ", "), "\n")
  }
  
  return(issues)
}

# Run cross-domain check
cross_result <- check_cross_domain(adsl_qc, adae_qc)
```

## QC Summary Report

```{r}
#| label: qc-summary

generate_qc_summary <- function(dup_result, missing_result, ct_result, date_result, age_result, cross_result) {
  cat("\n" , "="*60, "\n")
  cat("            QUALITY CONTROL SUMMARY REPORT\n")
  cat("="*60, "\n")
  cat("Dataset: ADSL\n")
  cat("QC Date:", format(Sys.Date(), "%Y-%m-%d"), "\n")
  cat("Total Records:", nrow(adsl_qc), "\n")
  cat("-"*60, "\n")
  
  # Summary of issues
  total_issues <- 0
  
  cat("ISSUE SUMMARY:\n")
  
  if (dup_result$status == "FAIL") {
    cat("• Duplicate subjects:", dup_result$count, "\n")
    total_issues <- total_issues + dup_result$count
  }
  
  if (length(missing_result) > 0) {
    missing_count <- sum(sapply(missing_result, function(x) x$count))
    cat("• Missing critical values:", missing_count, "\n")
    total_issues <- total_issues + missing_count
  }
  
  if (length(ct_result) > 0) {
    ct_count <- sum(sapply(ct_result, nrow))
    cat("• Controlled terminology issues:", ct_count, "\n")
    total_issues <- total_issues + ct_count
  }
  
  if (length(date_result) > 0) {
    date_count <- sum(sapply(date_result, nrow))
    cat("• Date logic issues:", date_count, "\n")
    total_issues <- total_issues + date_count
  }
  
  if (length(cross_result) > 0) {
    cross_count <- length(unlist(cross_result))
    cat("• Cross-domain issues:", cross_count, "\n")
    total_issues <- total_issues + cross_count
  }
  
  cat("-"*60, "\n")
  cat("TOTAL ISSUES IDENTIFIED:", total_issues, "\n")
  
  if (total_issues == 0) {
    cat("🎉 QC STATUS: PASS - No issues found\n")
  } else {
    cat("⚠️  QC STATUS: REVIEW REQUIRED\n")
  }
  
  cat("="*60, "\n")
}

# Generate summary report
generate_qc_summary(dup_result, missing_result, ct_result, date_result, age_result, cross_result)
```

## AI-Assisted QC with GitHub Copilot

### Copilot Prompts for QC

Here are effective GitHub Copilot prompts for clinical data QC:

```{r}
#| label: copilot-prompts
#| eval: false

# Example Copilot prompts for QC programming:

# 1. Generate validation function
# Prompt: "Create a function to validate SDTM dates against FDA requirements"

# 2. Check data completeness  
# Prompt: "Write R code to check missing values in clinical trial dataset by treatment group"

# 3. Controlled terminology validation
# Prompt: "Create validation rules for CDISC controlled terminology in R using dplyr"

# 4. Cross-domain referential integrity
# Prompt: "Write function to check that all subjects in AE domain exist in DM domain"

# 5. Statistical outlier detection
# Prompt: "Identify statistical outliers in vital signs data using R"
```

### Advanced QC Techniques

```{r}
#| label: advanced-qc

# Statistical outlier detection for continuous variables
detect_outliers <- function(data, var_name, method = "iqr") {
  cat(paste("Outlier detection for", var_name, "using", method, "method:\n"))
  
  values <- data[[var_name]][!is.na(data[[var_name]])]
  
  if (method == "iqr") {
    Q1 <- quantile(values, 0.25)
    Q3 <- quantile(values, 0.75)
    IQR <- Q3 - Q1
    lower_bound <- Q1 - 1.5 * IQR
    upper_bound <- Q3 + 1.5 * IQR
    
    outliers <- data %>%
      filter(!is.na(!!sym(var_name)) & 
             (!!sym(var_name) < lower_bound | !!sym(var_name) > upper_bound)) %>%
      select(USUBJID, !!sym(var_name))
    
    if (nrow(outliers) > 0) {
      cat("Outliers detected:\n")
      print(outliers)
    } else {
      cat("No outliers detected\n")
    }
  }
  
  return(outliers)
}

# Check for age outliers
age_outliers <- detect_outliers(adsl_qc, "AGE")
```

## Best Practices

### QC Checklist

1. **Data Completeness**
   - ✅ Check for missing key identifier variables
   - ✅ Validate completeness of critical efficacy/safety variables
   - ✅ Assess missing data patterns by treatment group

2. **Data Consistency** 
   - ✅ Validate controlled terminology adherence
   - ✅ Check date logic and chronological order
   - ✅ Verify cross-domain referential integrity

3. **Statistical Validation**
   - ✅ Identify statistical outliers
   - ✅ Check for impossible/implausible values
   - ✅ Validate derived variables calculations

4. **Regulatory Compliance**
   - ✅ CDISC standards compliance
   - ✅ FDA submission requirements
   - ✅ ICH-GCP guidelines adherence

### GitHub Copilot Integration Tips

- Use descriptive variable names to help Copilot understand context
- Write clear comments explaining the clinical logic
- Break complex validations into smaller, focused functions
- Use domain-specific terminology (CDISC, ICH, FDA) in prompts

---

*This QC toolkit provides a comprehensive framework for clinical data validation using modern R programming techniques with AI assistance.*